Hierarchical wrinkling in a confined permeable biogel

### 2015-02-14 fig_plis/article_plis.tex
Abstract minimise, characterise (en_GB) K\'arm\'an water -> solvent Ccl: pas court vs long mais proche/loin des murs

\begin{abstract}
The slow acidification of caseins, a family of milk proteins, is well known to result in the formation of gels -- this is indeed one of the key steps to make yoghurt. Here, using the casein proteins as  building blocks and the acidulent glucono-$\delta$-lactone, we design model thin biogel layers that wrinkles in cascade. Using a combination of titration, rheology, light microscopy and confocal microscopy, we demonstrate that, during the acidification, the gel shrinks and swells creating constrains that induce Darcy and Poiseuille flows and wrinkles the gel. We develop an elasto-hydrodynamic model that predicts correctly the wrinkling wavelength. Our results should be universal for acid-induced protein gels as it is based on pH induced charge stabilisation/destabilisation and on control of boundary conditions. As such it could set a benchmark to gain fundamental insights into wrinkled biological tissues, to texture food or to design surfaces for optical purposes.
\end{abstract}

@@

There are many ways and reasons for a film to wrinkle. An elastic film buckles due to excess area with respect to its boundaries and wrinkles when buckling is hindered by a substrate. The selection of the preferred wrinkling wavelength is peculiar to the situation and is related to the mismatch of elastic properties between the surface and the substrate~\cite{Gough1940, Bijlaard1946, Biot1957, Bowden1998, Cerda2003, Genzer2006}. On the biology side wrinkling-controlled morphogenesis is ubiquitous. Ageing and the loss of elastic fibres makes our skin wrinkles~\cite{Bissett1987, Genzer2006}. Difference in growth rates between the gut tube and its dorsal anchoring is responsible for the vilification of guts~\cite{Savin2011, Ciarletta2014, Shyer2013}. Localized cell death in biofilms focuses mechanical forces and initiates 3D labyrinth pattern~\cite{Trejo2013, Asally2012}. On the physics side the last two decades have seen a bloom of methods to obtain well-controlled patterns via linear~\cite{Bowden1998, Genzer2006, Hu1998, Kim2010, Vandeparre2011, Li2013} or nonlinear wrinkling~\cite{Efimenko2005, Guvendiren2010, Kim2011a, Brau2011}. Such patterns can be triggered by temperature dilation~\cite{Bowden1998}, swelling~\cite{Hu1998,Kim2010} or the removal of pre-strain~\cite{Genzer2006}.

@@

\begin{figure}[ht!]
	\includegraphics{dynamics}%
	\caption{Dynamics of pattern formation in a confined film of casein gel (caseinate 4\%w, GDL 4\%w). (a) Light transmission microscopy. Successive generations of patterns are highlighted in colour to stress the absence of growth after formation. (b) 3D reconstruction from fluorescent confocal microscopy (see Supplementary Table~\ref{tab:data}c). Scale bars are \SI{1}{\milli\metre}.}%
	\label{fig:dynamics}%
\end{figure}

@@

The stability of caseins dispersion strongly depend on the pH of the solution. Supplementary Video 3 shows the titration of a caseins dispersion at 1\%w by 1 molar HCl. At pH=7, the caseins are well dispersed then then flocculate between pH=5.5 to 4 and finally at pH=3 the solution become clear again. However as one increases the caseins concentration up to 4\%w, at low pH the caseins redispersion is not total, Supplementary Video 4. The state diagram of caseins as a function of the pH is related to charge stabilization. At neutral pH caseins are charged negatively and repel each other. As the pH decreases to the isoelectric point of the caseins (pH$\approx 4.6$), caseins molecules loose their negative net charge and tend to stick to each other~\cite{Roefs1986,Bremer1989,Lucey1998,Dickinson2002}. Beyond the isoelectric point casein acquire a positive net charge and therefore partially re-dissolve. In the inset of Figure~\ref{fig:acidification}b, we quantify this effect and measure with UV absorption spectroscopy the amount of free caseins within the 4\%w dispersion as a function of pH (see Supplementary Methods). Starting at pH=6.8 all the casein molecules are well dispersed and the free casein fraction $x_\text{free}$ is 1. Upon decreasing the pH of the solution to the isolectric point the free casein fraction drop to almost 0, almost all the casein molecule are aggregated. Further decrease of the pH of the solution lead to a partial redissolution of the casein aggregates. At pH=2 the fraction of free casein is 0.35.

@@

To form a space spanning gel, flocculation needs to be prevented. We use of glucono-$\delta$-lactone (GDL) a molecule  that slowly hydrolyses into an acid and thus lowers slowly, continuously and homogeneously the pH of the medium, Figure~\ref{fig:acidification}a. This GDL controlled acidification enables the formation of a bulk gel phase below pH=5.3 with $x_\text{free}\approx 0$. As shown on Figure~\ref{fig:acidification}a and b, when adding enough GDL the pH lowers beyond the isoeletric point and $x_\text{free}$ rises again. We monitor the GDL induced gelation with a rheometer in cone plate geometry. As casein strongly adsorbs to surfaces, the strain and stress are well transmitted to the casein dispersion and we can measure the elastic and viscous modulus of the casein dispersion as a function of time, Figure~\ref{fig:acidification}c. The mechanical properties of the gel are closely related to the fraction of free casein. First, the dispersion solidifies then reaches its maximum elasticity (500 Pa) at the isoelectric point and finally the elastic modulus decreases down to 100 Pa. This finite value indicates that the sample is still a gel.

@@

Figure~\ref{fig:sideview}a shows confocal measurements in the case where both top and bottom walls are coated, adhesion turned off, a configuration favourable to wrinkling (see also Supplementary Videos 7 and 8). Initially the dispersion is homogeneous and fills the entire slit. The black margins correspond to the glass walls. Around \SI{23}{\minute}, the gel forms and simultaneously becomes thinner and more concentrated - the fluorescence signal which is proportional to the casein concentration is brighter compared to the initial situation. This effect is the fingerprint of syn\ae{}resis: as gelation proceeds, the solvent is expelled on both the upper and lower sides. We describe this situation as a porous elastic film of thickness $h$ sandwiched between two viscous layers of respective thickness $H_1$ and $H_2$ in a cell of thickness $e=H_1+h+H_2$ (Figure~\ref{fig:sideview}b). Extracting the top and bottom surface of the gel phase allows for quantitative measurements of the relative volume (Figure~\ref{fig:sideview}c) and excess surface area of the gel film (Figure~\ref{fig:sideview}d) (see Supplementary Methods). As expected from the system design, the evolution of the volume is non-monotonous. After reaching a minimum around 35\% of its initial value at \SI{23}{\minute}, the volume increases. Shortly after,  at \SI{36}{\minute}, the planar film destabilizes and starts to wrinkle in the $z$-direction creating excess area. This confirms that the instability is due to swelling. Around \SI{38}{\minute}, the amplitude of the wrinkles reach the thickness of the slit. The size of the primary pattern is now settled, Supplementary Figure~\ref{fig:growth}. However swelling continues, the gel flattens on both top and bottom walls (\SI{44}{\minute}) and finally buckles back (\SI{53}{\minute}) to form the secondary pattern (\SI{1}{\hour}21). When the excess volume saturates, the dynamics stops. The amount of GDL controls the ability of the gel to swell: at low GDL concentrations we only observe the primary pattern, at high GDL concentrations we observe the formation of higher order patterns.

Note that adhesion to side walls is necessary to prevent total syn\ae{}resis and thus allows subsequent planar stresses to develop while swelling: the only situation where we observe wrinkling is when the adhesion on both top and bottom wall is turned off. We also confirmed that at any given point in time and space the thickness $h$ of the film of gel is homogeneous. The contrast in transmission microscopy (Figure~\ref{fig:dynamics}a) is therefore not due to thickness inhomogeneities but to altitude gradients. Indeed the length of the optical path along the vertical direction $z$ through the gel phase is larger where the gel makes a slope compared to a flat situation, as indicated by the yellow vertical lines in Figure~\ref{fig:sideview}b.

@@

The determination of the characteristic length $\lambda$ of a wrinkling pattern is an old problem. Large deformations of thin sheets is governed by F\"{o}ppl-von Karman equations~\cite{landau1986}, essentially impossible to solve in analytical form. However using scaling and asymptotic arguments one can obtain the dominant wavelength of the wrinkling pattern and isolate the physically meaningful ingredients~\cite{Cerda2003}. An elastic film buckles due to excess area with respect to its boundaries. Wrinkling is a buckling hindered by a substrate: the mismatch of elastic properties between the film and the substrate selects the preferred wrinkling wavelength~\cite{Gough1940, Bijlaard1946}. $\lambda$ is then set by the competition between the flexural modulus of the film $B=E/(1-\nu^2)$ ($E$ Young modulus, $\nu$ Poisson ratio) and a transverse load $\sigma_\perp$. For small deflections of  amplitude $A$, $\lambda$ scales as~\cite{Vella2009,Kolinski2009}: 
%
\begin{equation}
\lambda \sim \left( A h^3 \frac{B}{\sigma_{\perp}} \right)^{1/4}.
\label{eq:lstar}
\end{equation}

@@

In our situation we now have to determine $\sigma_\perp$. We identify three possible candidates for the transverse load $\sigma_\perp$: the weight or buoyancy of the gel, the porous flow through the gel or the viscous flows in the surrounding water layers. We confirmed that gravity plays no significant role by successfully repeating the wrinkling pattern in a cell held vertically, i.e. with the weight acting longitudinally rather than transversely.

@@

We consider two limit scenarios that leads to wrinkling patterns. In a first limit, the film of gel sits without sticking to the bottom wall of the cell, i.e. $H_1\rightarrow0$, see Supplementary Video 7. Creating a 2D blister requires flowing solvent of viscosity $\eta$ through the gel of permeability $\alpha$. Darcy law~\cite{Darcy1856} relates the pressure gradient, estimated as $\sigma_{\perp}/h$, to the volume flux per unit area $v = \frac{\alpha}{\eta}\frac{\sigma_{\perp}}{h} = \frac{dA}{dt}$. Injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Darcy mode as
\begin{equation}
\lambda_D \sim h^{1/2} \alpha^{1/4} \Upsilon^{1/4}
\end{equation}
where $\Upsilon = B\tau/\eta$ is a dimensionless viscoelastic factor quantifying the relative stiffness of the film and the solvent at a characteristic time $\tau$. This expression emphasizes the kinetic nature of the wavelength selection through Darcy flow. However we note that the wavelength of a generation of the pattern does not evolve with time (see Figure~\ref{fig:dynamics}a and Supplementary Figure~\ref{fig:growth}) once the gel is touching both walls. This arrest may be due to bending rigidity~\cite{LeGoff2014}. Here the relevant time scale $\tau$ corresponds to the times it takes for the wrinkle amplitude to saturate (see Supplementary Method for the precise definition).

The opposite limiting case is when the gel film lies in the middle of the cell ($H_1=H_2$, see Supplementary Video 8) and impermeable ($\alpha\rightarrow 0$). Here destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the transverse load across the gel $\sigma_{\perp}$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim \sigma_{\perp} \frac{H^3}{\eta\lambda^2}$~\cite{Poiseuille1842}. Injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognize the scaling for the wrinkles of an elastic film on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural modulus of the film and that of the substrate. Indeed within the lubrication approximation or at low Reynolds numbers one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.

@@

Darcy and Poiseuille models yield wavelengths that are of the same order in most of the cases under study. Thus only systematic measurements of the experimental $\lambda$ together with estimates for all the parameters involved in either model shall allow one to discriminate between them (see Supplementary Method for details on such measurements). As summarized in Supplementary Table~\ref{tab:data} we systematically vary (i) the thickness of the cell, impacting $h$ and the maximal amplitude, (ii) the gel composition and thus both its stiffness and its permeability, (iii) the viscosity of the solvent by adding glycerol, which also impacted the gel stiffness, permeability and $\tau$. Finally, the only parameter that we have no control upon is the initial altitude $H$ of the gel before wrinkling that varies randomly. However we are able to obtain more data at small $H/H^*$ ratios by considering not only the primary patterns but also the secondary ones. Since the saturation of the primary pattern flattens the gel on the top and bottom walls, the secondary blisters always appear in a Darcy situation ($H\ll H^*$). Here the wavelength associated to the secondary blister is taken as half of that of the primary pattern as shown in Supplementary Figure~\ref{fig:generations}.

Our results are displayed in Figure~\ref{fig:DarcyPoiseuille}. Although they provide the correct order of magnitude for $\lambda$, neither model alone is able to quantitatively reproduce the full data set. However considering only the points where $H<H^*$ (respectively $H>H^*$), the Darcy (resp. Poiseuille) model fits the data well albeit with a prefactor. Note that it would be easy to miss out on the failure of the Poiseuille model at small wavelengths by adding an unexplained offset (dashed line in Figure~\ref{fig:DarcyPoiseuille}b). Nonetheless, a complete model (derived in Supplementary Methods), summing up the influence of both Poiseuille and Darcy dissipation shows a better agreement for all values of $H/H^*$.

@@

In conclusion we have taken advantage of the properties of the protein casein and carefully engineered the route to form permeable biogels that, in slit geometry, produce original cascade of wrinkling patterns when casein adhesion to the slit walls is turned off. We developed a model that captures the dynamical origin of the constrains exerted on the gel and predict the wrinkling wavelength: Poiseuille flows of the solvent above and below the gel layer for long wavelengths and a new mechanism, Darcy flow of the solvent through the gel layer for short wavelengths. Such a model experiment, because it relies on pH induced charge stabilisation/destabilisation, should be generalisable to different sort of proteins. As such it illustrates daily life problems such as wrinkles encountered while putting up wall paper and could set a benchmark to explore potential applications in micro-fingerprinting~\cite{Bae2015} or in optical devices such as diffraction gratings or Fresnel lenses~\cite{Li2013}, and the possibility of wrinkling in confined porous soft materials immersed in a buoyancy-matched viscous medium such as biological tissues~\cite{Ma2004,Longley2013}.

@@

Home made optical cells are composed of a slide (RS France) and a cover slip (Menzel-Gl\"aser) spaced by parafilm (Parafilm). The optical cell is briefly heated on a hot plate so that the parafilm welds to the slide and the cover slip. Gently applying a pressure on the cover slip while being heated allows us to tune the thickness of the cell from \SIrange{50}{150}{\micro\metre}. The slit is filled with the casein dispersion, immediately sealed using ultraviolet-cured glue (Norland Optical) and placed under the microscope before gelation could take place. To prevent adsorption of caseins to surfaces, microscope slides and cover slips used to make microscopy cells are cleaned and subsequently coated with a silane agent (3-(trimethoxysilyl) propyl methacrylate, Sigma Aldrich) from which a polyacrylamide brush is polymerized from methacrylate groups (Euromedex).

@@

Minimizing $\sigma_\parallel$ as a function of $q$ yields $hq = \left(6B_s/B\right)^{1/3}$ \cite{Biot1957,Cerda2003} so that the dominant wavelength is 
\begin{equation}
\lambda = 2\pi h \left(\frac{B}{6B_s}\right)^{1/3}.
\label{eq:lambdaElEl}
\end{equation}

@@

Combining Equations~(\ref{eq:mecheqop}) and (\ref{eq:PoiseuilleLoad}), the compression load writes
\begin{equation}
\sigma_\parallel = \frac{1}{12}B h^2 q^2 + 12\frac{\eta}{\tau}\frac{1}{H^3 h q^4}
\label{eq:sigma0P}
\end{equation}
and is minimized at $q_P$ such that
\begin{equation}
q_P^6 = 2\times 12^2 \frac{1}{h^3H^3\Upsilon},
\label{eq:qP}
\end{equation}
with $\Upsilon = B\tau/\eta$ the viscoelastic factor. The dominant wavelength for the Poiseuille model is thus
\begin{equation}
\lambda_P = \pi\sqrt{hH}\left(\frac{2}{9}\Upsilon\right)^{1/6}
\end{equation}

@@

Combining equations~(\ref{eq:mecheqop}) and (\ref{eq:DarcyLoad}), the compression load writes
\begin{equation}
\sigma_\parallel = \frac{1}{12}B h^2 q^2 + \frac{\eta}{\tau}\frac{1}{\alpha q^2}
\label{eq:sigma0D}
\end{equation}
and is minimized at $q_D$ such that
\begin{equation}
q_D^4 = \frac{12}{h^2\alpha\Upsilon}.
\label{eq:qD}
\end{equation}

@@

Analytic minimization of $\sigma_\parallel$ knowing $\sigma_\perp$ is possible but cumbersome. Instead we rewrite equation~(\ref{eq:mecheqop}) in terms of previously calculated $\lambda_P$ and $\lambda_D$:
\begin{align}
\frac{12\sigma_\parallel}{q_D^2 h^2 B} &= \left(\frac{q}{q_D}\right)^2 + \frac{1}{2\frac{q_D^2}{q_P^6}q^4  + \left(\frac{q}{q_D}\right)^2}\\
&= \left(\frac{\lambda_D}{\lambda}\right)^2 + \frac{1}{2\frac{\lambda_P^6}{\lambda_D^2\lambda^4}  + \left(\frac{\lambda_D}{\lambda}\right)^2}
\end{align}
and we minimize this expression numerically.

\subsection*{Samples characteristics}
To test the theoretical predictions we have fully characterized 9 samples, see Supplementary Table \ref{tab:data}.

>>>>

\begin{abstract}
The slow acidification of caseins, a family of milk proteins, is well known to result in the formation of gels, e.g. yoghurt. Less known is the partial redissolution of the gel at even lower pH. Here, using caseins as pH-responsive bulding blocks and the acidulent glucono-$\delta$-lactone, we design a model biogel film that spontaneously wrinkles in cascade, provided non-adhesive confinement. Using a combination of titration, rheology, light microscopy and confocal microscopy, we demonstrate that, during continuous acidification, the gel first shrinks and then swells, inducing winkling. Taking into account both Darcy flow though the gel and Poiseuille flow in the surrounding solvent, we develop a model that predicts correctly the wrinkling wavelength. Our results should be universal for acid-induced protein gels as it is based on pH-induced charge stabilisation/destabilisation. As such it could set a benchmark to gain fundamental insights into wrinkled biological tissues, to texture food or to design surfaces for optical purposes.
\end{abstract}

@@

There are many ways and reasons for a film to wrinkle. An elastic film buckles due to excess area with respect to its boundaries and wrinkles when buckling is hindered by a substrate. The selection of the preferred wrinkling wavelength is peculiar to the situation and is related to the mismatch of elastic properties between the surface and the substrate~\cite{Gough1940, Bijlaard1946, Biot1957, Bowden1998, Cerda2003, Genzer2006}. On the biology side wrinkling-controlled morphogenesis is ubiquitous. Ageing and the loss of elastic fibres makes our skin wrinkle~\cite{Bissett1987, Genzer2006}. Difference in growth rates between the gut tube and its dorsal anchoring is responsible for the vilification of guts~\cite{Savin2011, Ciarletta2014, Shyer2013}. Localised cell death in biofilms focuses mechanical forces and initiates 3D labyrinth pattern~\cite{Trejo2013, Asally2012}. On the physics side the last two decades have seen a bloom of methods to obtain well-controlled patterns via linear~\cite{Bowden1998, Genzer2006, Hu1998, Kim2010, Vandeparre2011, Li2013} or nonlinear wrinkling~\cite{Efimenko2005, Guvendiren2010, Kim2011a, Brau2011}. Such patterns can be triggered by temperature dilation~\cite{Bowden1998}, swelling~\cite{Hu1998,Kim2010} or the removal of pre-strain~\cite{Genzer2006}.

@@

\begin{figure}
	\includegraphics{dynamics}%
	\caption{Dynamics of pattern formation in a confined film of casein gel (caseinate 4\%w, GDL 4\%w). (a) Light transmission microscopy. Successive generations of patterns are highlighted in colour to stress the absence of growth after formation. (b) 3D reconstruction from fluorescent confocal microscopy (see Supplementary Table~\ref{tab:data}c). Scale bars are \SI{1}{\milli\metre}.}%
	\label{fig:dynamics}%
\end{figure}

@@

The stability of caseins dispersion strongly depend on the pH of the solution. Supplementary Video 3 shows the titration of a caseins dispersion at 1\%w by 1 molar HCl. At pH=7, the caseins are well dispersed then then flocculate between pH=5.5 to 4 and finally at pH=3 the solution become clear again. However as one increases the caseins concentration up to 4\%w, at low pH the caseins redispersion is not total, Supplementary Video 4. The state diagram of caseins as a function of the pH is related to charge stabilisation. At neutral pH caseins are charged negatively and repel each other. As the pH decreases to the isoelectric point of the caseins (pH$\approx 4.6$), casein molecules loose their negative net charge and tend to stick to each other~\cite{Roefs1986,Bremer1989,Lucey1998,Dickinson2002}. Beyond the isoelectric point caseins acquire a positive net charge and therefore partially re-dissolve. In the inset of Figure~\ref{fig:acidification}b, we quantify this effect and measure with UV absorption spectroscopy the amount of free caseins within the 4\%w dispersion as a function of pH (see Supplementary Methods). Starting at pH=6.8 all the casein molecules are well dispersed and the free casein fraction $x_\text{free}$ is 1. Upon decreasing the pH of the solution to the isolectric point the free casein fraction drops to almost 0, almost all the casein molecule are aggregated. Further decrease of the pH of the solution lead to a partial redissolution of the casein aggregates. At pH=2 the fraction of free casein is 0.35.

@@

To form a space spanning gel, flocculation needs to be prevented. We use glucono-$\delta$-lactone (GDL) a molecule  that slowly hydrolyses into an acid and thus lowers slowly, continuously and homogeneously the pH of the medium, Figure~\ref{fig:acidification}a. This GDL controlled acidification enables the formation of a bulk gel phase below pH=5.3 with $x_\text{free}\approx 0$. As shown on Figure~\ref{fig:acidification}a and b, when adding enough GDL the pH lowers beyond the isoeletric point and $x_\text{free}$ rises again. We monitor the GDL induced gelation with a rheometer in cone plate geometry. As casein strongly adsorbs to surfaces, the strain and stress are well transmitted to the casein dispersion and we can measure the elastic and viscous modulus of the casein dispersion as a function of time, Figure~\ref{fig:acidification}c. The mechanical properties of the gel are closely related to the fraction of free casein. First, the dispersion solidifies then reaches its maximum elasticity (500 Pa) at the isoelectric point and finally the elastic modulus decreases down to 100 Pa. This finite value indicates that the sample is still a gel.

@@

Figure~\ref{fig:sideview}a shows confocal measurements in the case where both top and bottom walls are coated, adhesion turned off, a configuration favourable to wrinkling (see also Supplementary Videos 7 and 8). Initially the dispersion is homogeneous and fills the entire slit. The black margins correspond to the glass walls. Around \SI{23}{\minute}, the gel forms and simultaneously becomes thinner and more concentrated - the fluorescence signal which is proportional to the casein concentration is brighter compared to the initial situation. This effect is the fingerprint of syn\ae{}resis: as gelation proceeds, the solvent is expelled on both the upper and lower sides. We describe this situation as a porous elastic film of thickness $h$ sandwiched between two viscous layers of respective thickness $H_1$ and $H_2$ in a cell of thickness $e=H_1+h+H_2$ (Figure~\ref{fig:sideview}b). Extracting the top and bottom surface of the gel phase allows for quantitative measurements of the relative volume (Figure~\ref{fig:sideview}c) and excess surface area of the gel film (Figure~\ref{fig:sideview}d) (see Supplementary Methods). As expected from the system design, the evolution of the volume is non-monotonous. After reaching a minimum around 35\% of its initial value at \SI{23}{\minute}, the volume increases. Shortly after,  at \SI{36}{\minute}, the planar film destabilises and starts to wrinkle in the $z$-direction creating excess area. This confirms that the instability is due to swelling. Around \SI{38}{\minute}, the amplitude of the wrinkles reach the thickness of the slit. The size of the primary pattern is now settled, Supplementary Figure~\ref{fig:growth}. However swelling continues, the gel flattens on both top and bottom walls (\SI{44}{\minute}) and finally buckles back (\SI{53}{\minute}) to form the secondary pattern (\SI{1}{\hour}21). When the excess volume saturates, the dynamics stops. The amount of GDL controls the ability of the gel to swell: at low GDL concentrations we only observe the primary pattern, at high GDL concentrations we observe the formation of higher order patterns.

Note that adhesion to side walls is necessary to prevent total syn\ae{}resis and thus allows subsequent planar stresses to develop while swelling: the only situation where we observe wrinkling is when the adhesion on both top and bottom wall is turned off. We also confirm that at any given point in time and space the thickness $h$ of the film of gel is homogeneous. The contrast in transmission microscopy (Figure~\ref{fig:dynamics}a) is therefore not due to thickness inhomogeneities but to altitude gradients. Indeed the length of the optical path along the vertical direction $z$ through the gel phase is larger where the gel makes a slope compared to a flat situation, as indicated by the yellow vertical lines in Figure~\ref{fig:sideview}b.

@@

The determination of the characteristic length $\lambda$ of a wrinkling pattern is an old problem. Large deformations of thin sheets is governed by F\"{o}ppl-von K\'arm\'an equations~\cite{landau1986}, essentially impossible to solve in analytical form. However using scaling and asymptotic arguments one can obtain the dominant wavelength of the wrinkling pattern and isolate the physically meaningful ingredients~\cite{Cerda2003}. An elastic film buckles due to excess area with respect to its boundaries. Wrinkling is a buckling hindered by a substrate: the mismatch of elastic properties between the film and the substrate selects the preferred wrinkling wavelength~\cite{Gough1940, Bijlaard1946}. $\lambda$ is then set by the competition between the flexural modulus of the film $B=E/(1-\nu^2)$ ($E$ Young modulus, $\nu$ Poisson ratio) and a transverse load $\sigma_\perp$. For small deflections of  amplitude $A$, $\lambda$ scales as~\cite{Vella2009,Kolinski2009}: 
%
\begin{equation}
\lambda \sim \left( A h^3 \frac{B}{\sigma_{\perp}} \right)^{1/4}.
\label{eq:lstar}
\end{equation}

@@

In our situation we now have to determine $\sigma_\perp$. We identify three possible candidates for the transverse load $\sigma_\perp$: the weight or buoyancy of the gel, the porous flow through the gel or the viscous flows in the surrounding solvent layers. We confirmed that gravity plays no significant role by successfully repeating the wrinkling pattern in a cell held vertically, i.e. with the weight acting longitudinally rather than transversely.

@@

We consider two limit scenarios that leads to wrinkling patterns. In a first limit, the film of gel sits without sticking to the bottom wall of the cell, i.e. $H_1\rightarrow0$, see Supplementary Video 7. Creating a 2D blister requires flowing solvent of viscosity $\eta$ through the gel of permeability $\alpha$. Darcy law~\cite{Darcy1856} relates the pressure gradient, estimated as $\sigma_{\perp}/h$, to the volume flux per unit area $v = \frac{\alpha}{\eta}\frac{\sigma_{\perp}}{h} = \frac{dA}{dt}$. Injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Darcy mode as
\begin{equation}
\lambda_D \sim h^{1/2} \alpha^{1/4} \Upsilon^{1/4}
\end{equation}
where $\Upsilon = B\tau/\eta$ is a dimensionless viscoelastic factor quantifying the relative stiffness of the film and the solvent at a characteristic time $\tau$. This expression emphasises the kinetic nature of the wavelength selection through Darcy flow. However we note that the wavelength of a generation of the pattern does not evolve with time once the gel is touching both walls(see Figure~\ref{fig:dynamics}a and Supplementary Figure~\ref{fig:growth}). This arrest may be due to bending rigidity~\cite{LeGoff2014}. Here the relevant time scale $\tau$ corresponds to the times it takes for the wrinkle amplitude to saturate (see Supplementary Method for the precise definition).

The opposite limiting case is when the gel film lies in the middle of the cell ($H_1=H_2$, see Supplementary Video 8) and impermeable ($\alpha\rightarrow 0$). Here destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the transverse load across the gel $\sigma_{\perp}$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim \sigma_{\perp} \frac{H^3}{\eta\lambda^2}$~\cite{Poiseuille1842}. Injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognise the scaling for the wrinkles of an elastic film on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural modulus of the film and that of the substrate. Indeed within the lubrication approximation or at low Reynolds numbers one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.

@@

Darcy and Poiseuille models yield wavelengths that are of the same order in most of the cases under study. Thus only systematic measurements of the experimental $\lambda$ together with estimates for all the parameters involved in either model shall allow one to discriminate between them (see Supplementary Method on such measurements). As summarised in Supplementary Table~\ref{tab:data} we systematically vary (i) the thickness of the cell, impacting $h$ and the maximal amplitude, (ii) the gel composition and thus both its stiffness and its permeability, (iii) the viscosity of the solvent by adding glycerol, which also impacted the gel stiffness, permeability and $\tau$. Finally, the only parameter that we have no control upon is the initial altitude $H$ of the gel before wrinkling that varies randomly. However we are able to obtain more data at small $H/H^*$ ratios by considering not only the primary patterns but also the secondary ones. Since the saturation of the primary pattern flattens the gel on the top and bottom walls, the secondary blisters always appear in a Darcy situation ($H\ll H^*$). Here the wavelength associated to the secondary blister is taken as half of that of the primary pattern as shown in Supplementary Figure~\ref{fig:generations}.

Our results are displayed in Figure~\ref{fig:DarcyPoiseuille}. Although they provide the correct order of magnitude for $\lambda$, neither model alone is able to quantitatively reproduce the full data set. However considering only the points where $H<H^*$ (respectively $H>H^*$), the Darcy (resp. Poiseuille) model fits the data well albeit with a prefactor. Note that it would be easy to miss out on the failure of the Poiseuille model at small wavelengths by adding an unexplained offset (dashed line in Figure~\ref{fig:DarcyPoiseuille}b). Nonetheless, a complete model (derived in Supplementary Methods), summing up the influence of both Poiseuille and Darcy dissipation shows a better agreement for the whole $H/H^*$ range.

@@

In conclusion we have taken advantage of the properties of the protein casein and carefully engineered the route to form permeable biogels that, in slit geometry, produce original cascade of wrinkling patterns when casein adhesion to the slit walls is turned off. We developed a model that captures the dynamical origin of the constrains exerted on the gel and predict the wrinkling wavelength: Poiseuille flows of the solvent above and below the gel when the film is far from any wall, Darcy flow of the solvent through the gel when the film is close to a wall. Such a model experiment, because it relies on pH induced charge stabilisation/destabilisation, should be generalisable to different sort of proteins. As such it illustrates daily life problems such as wrinkles encountered while putting up wall paper and could set a benchmark to explore potential applications in micro-fingerprinting~\cite{Bae2015} or in optical devices such as diffraction gratings or Fresnel lenses~\cite{Li2013}, and the possibility of wrinkling in confined porous soft materials immersed in a buoyancy-matched viscous medium such as biological tissues~\cite{Ma2004,Longley2013}.

@@

Home made optical cells are composed of a slide (RS France) and a cover slip (Menzel-Gl\"aser) spaced by parafilm (Parafilm). The optical cell is briefly heated on a hot plate so that the parafilm welds to the slide and the cover slip. Gently applying a pressure on the cover slip while being heated allows us to tune the thickness of the cell from \SIrange{50}{150}{\micro\metre}. The slit is filled with the casein dispersion, immediately sealed using ultraviolet-cured glue (Norland Optical) and placed under the microscope before gelation could take place. To prevent adsorption of caseins to surfaces, microscope slides and cover slips used to make microscopy cells are cleaned and subsequently coated with a silane agent (3-(trimethoxysilyl) propyl methacrylate, Sigma Aldrich) from which a polyacrylamide brush is polymerised from methacrylate groups (Euromedex).

@@

Minimising $\sigma_\parallel$ as a function of $q$ yields $hq = \left(6B_s/B\right)^{1/3}$ \cite{Biot1957,Cerda2003} so that the dominant wavelength is 
\begin{equation}
\lambda = 2\pi h \left(\frac{B}{6B_s}\right)^{1/3}.
\label{eq:lambdaElEl}
\end{equation}

@@

Combining Equations~(\ref{eq:mecheqop}) and (\ref{eq:PoiseuilleLoad}), the compression load writes
\begin{equation}
\sigma_\parallel = \frac{1}{12}B h^2 q^2 + 12\frac{\eta}{\tau}\frac{1}{H^3 h q^4}
\label{eq:sigma0P}
\end{equation}
and is minimised at $q_P$ such that
\begin{equation}
q_P^6 = 2\times 12^2 \frac{1}{h^3H^3\Upsilon},
\label{eq:qP}
\end{equation}
with $\Upsilon = B\tau/\eta$ the viscoelastic factor. The dominant wavelength for the Poiseuille model is thus
\begin{equation}
\lambda_P = \pi\sqrt{hH}\left(\frac{2}{9}\Upsilon\right)^{1/6}
\end{equation}

@@

Combining equations~(\ref{eq:mecheqop}) and (\ref{eq:DarcyLoad}), the compression load writes
\begin{equation}
\sigma_\parallel = \frac{1}{12}B h^2 q^2 + \frac{\eta}{\tau}\frac{1}{\alpha q^2}
\label{eq:sigma0D}
\end{equation}
and is minimised at $q_D$ such that
\begin{equation}
q_D^4 = \frac{12}{h^2\alpha\Upsilon}.
\label{eq:qD}
\end{equation}

@@

Analytic minimisation of $\sigma_\parallel$ knowing $\sigma_\perp$ is possible but cumbersome. Instead we rewrite equation~(\ref{eq:mecheqop}) in terms of previously calculated $\lambda_P$ and $\lambda_D$:
\begin{align}
\frac{12\sigma_\parallel}{q_D^2 h^2 B} &= \left(\frac{q}{q_D}\right)^2 + \frac{1}{2\frac{q_D^2}{q_P^6}q^4  + \left(\frac{q}{q_D}\right)^2}\\
&= \left(\frac{\lambda_D}{\lambda}\right)^2 + \frac{1}{2\frac{\lambda_P^6}{\lambda_D^2\lambda^4}  + \left(\frac{\lambda_D}{\lambda}\right)^2}
\end{align}
and we minimise this expression numerically.

\subsection*{Samples characteristics}
To test the theoretical predictions we have fully characterised 9 samples, see Supplementary Table \ref{tab:data}.
